evidence of adaptation-independent time-warp spike timing noise correlations in cortical networks

### 2020-08-12 manuscript.tex
Great!

In early visual cortex, trial-to-trial spike count variability is correlated across the population and shown to be modulated by two global factors, which fluctuate the level of excitability from trial-to-trial \cite{lin2015nature}. The first of these is an additive factor, which is explained by the interaction of evoked activity with population-wide spontaneous activity \cite{arieli1996dynamics, scholvinck2015cortical}. The second is a multiplicative factor, which also modulates the trial-to-trial fluctuation in excitability across the population, and has been proposed to be caused by top-down feedback from higher cortices \cite{lin2015nature}. Fluctuations in spontaneous activity have also been shown to gate sensory-evoked responses \cite{luczak2013gating}.

Fluctuations in trial-to-trial excitability are referred to herein as trial-to-trial fluctuations in continuous cortical state \cite{scholvinck2015cortical}, which may be multi-dimensional and can be separated into short and long-timescale changes. In this paper, long-timescale fluctuations refer to large changes taking place over multiple trials caused by processes such as neural adaptation in response to repetitive stimulation or changes in the level of anaesthesia. Short-timescale fluctuations refer to random trial-to-trial fluctuations in excitability around the current general level of excitability.

@@

This study aims to characterise the non-trivial nature of trial-to-trial spike time (rather than spike count) noise correlations, as a component step towards investigating the modulation of spike times by short-timescale fluctuations of a low-dimensional continuous cortical state. 
% which would introduce the possibility of state-dependent encoding and decoding (see below). 
The characterisation controls for longer timescale changes in excitability to estimate the short-timescale state-modulated representation of a stimulus at a given point in time. 
% and a novel prediction of the encoding capacity under a state-dependent spike time encoding regime. 
Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.

@@

Responses following sensory stimulation can be analysed to explore such principles of neural coding. 
% Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
A multi-neuron response can be expressed as a point in space, where each dimension represents the value of a spike train metric. Responses from multiple presentations of a stimulus form a sample stimulus-response distribution in this space, which can be used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig. 1a).

@@

% Information theoretic approaches aim to understand 
If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a). Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.
`Noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimuli \cite{averbeck2006neural}. `Signal correlations' concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}).

@@

It is important to consider that if downstream neurons are also modulated by the same population-wide short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
When the short-timescale fluctuating state has a certain value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig. 1c). State-dependent decoding would therefore offer an improved encoding capacity, compared to the classical fixed-state decoder interpretation.

% Correlative structure between the responses of different neurons in single stimulus response distributions indicates the presence of trial-to-trial noise shared between the responses of multiple neurons. These are classically referred to as `noise correlations' \cite{averbeck2006neural}, rather than

@@

Characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}. 
% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
% Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also 
Long-term changes in cortical state, such as adpatation, could lead to correlative structure in the stimulus-response distribution (Fig. X), and must be controlled for when characterising short-timescale or `adaptation-independent' noise correlations.

@@

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of short-timescale spike time noise correlations for first-spike patterns, whilst controlling for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
% towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state. The classical `stimulus-dependency' of noise correlations is left to future studies. 
A previous study analysing the same data set found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}. Another study analysing the data set could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}. 
% The spike time noise correlations uncovered in this study 
% offer a new opportunity to study population-wide modulation by low-dimensional latent states and 
% controls for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
% and was confirmed to occur for 7 out of 8 neuron groups in this study.

@@

Interestingly, many of the correlated clusters had non-unitary correlation angles. 
Importantly, many clusters remained correlated with non-unitary gradient after controlling for longer-timescale neural adaptation,
% using time series analysis, 
indicative of short-timescale fluctuating state that affects different neurons differently.
% giving evidence of precise non-adaptive noise correlations
% Such non-unitary spike time correlation angles cannot be 
This cannot explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle).

@@

The characterisation of adaptation-independent clustered spike-timing noise correlations is useful under the fixed-state decoder interpretations where noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}. More interestingly, these noise correlations open the possibility of state-dependent spike time coding. Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaptation-controlled correlated cluster. This gave a prediction of the resolution of possible state-dependent encoding, which surpasses previously proposed coding forms.

Adaptation-independent non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers. This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner. 
% The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. 
Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by low trial overlap between cluster. The work of this paper 
% in extracting and characterising adaptation-independent correlated clusters provides 
a useful component step towards investigating this. 
% confirming or rejecting state-dependent spike-time coding. 
The results show that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.

>>>>

In early visual cortex, trial-to-trial spike count variability is correlated across the population and is modulated by two global factors, which fluctuate the level of excitability from trial-to-trial \cite{lin2015nature}. The first of these is an additive factor, which is explained by the interaction of evoked activity with population-wide spontaneous activity \cite{arieli1996dynamics, scholvinck2015cortical}. The second is a multiplicative factor, which also modulates the trial-to-trial fluctuation in excitability across the population, and has been proposed to be caused by top-down feedback from higher cortices \cite{lin2015nature}. Fluctuations in spontaneous activity have also been shown to gate sensory-evoked responses \cite{luczak2013gating}.

Fluctuations in trial-to-trial excitability are referred to herein as trial-to-trial fluctuations in continuous cortical state \cite{scholvinck2015cortical}, which may be multi-dimensional and can be separated into short and long-timescale changes. In this paper, long-timescale fluctuations refer to large changes taking place over multiple trials caused by processes such as neural adaptation in response to repetitive stimulation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}, or changes in the level of anaesthesia. Short-timescale fluctuations refer to random trial-to-trial fluctuations in excitability around the current general level of excitability.

@@

This study aims to characterise the non-trivial nature of trial-to-trial spike time (rather than spike count) noise correlations, as a component step towards investigating the modulation of spike times by short-timescale fluctuations of a low-dimensional continuous cortical state. 
% which would introduce the possibility of state-dependent encoding and decoding (see below). 
The characterisation controls for longer timescale changes in excitability to estimate how stimulus representations are modulated by trial-to-trial fluctuations in state at a given point in time. 
% This offers a novel prediction of the encoding capacity under a state-dependent spike time encoding regime. 
Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.

@@

Responses following sensory stimulation can be analysed to explore such principles of neural coding. 
% Due to response variability, characterising the neural representation of a single stimulus is a matter of characterising the \textit{distribution} of possible neural responses for the stimulus. 
A multi-neuron response can be expressed as a point in space, where each dimension represents a spike train metric. Responses to multiple stimulus presentations form a sample stimulus-response distribution in this space, which can be used to estimate an underlying response distribution from which single-trial responses are `drawn' (Fig. 1a).
% Information theoretic approaches aim to understand 
If two sensory stimuli have a cognitively relevant difference, downstream neurons must be able to respond differently to samples from the response distribution of either stimulus. This is possible if the response distributions are highly non-overlapping in dimensions that downstream neurons are sensitive to, such as relative spike time differences (\cite{gasparini2006state, branco2010dendritic, branco2011synaptic}; Fig. 1a).

@@

Neural coding classically considers such response-distributions with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}. `Noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimuli, whilst `signal correlations' concern tuning similarity between neurons. With respect to fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.

@@

It is important to consider that if downstream neurons are also modulated by the same population-wide short-timescale fluctuating state, then downstream `decoder' neurons could respond in a state-dependent manner.
When the fluctuating state has a value on a single trial, a response comes from a \textit{state-conditioned} stimulus-response distribution, which would occupy a smaller region of response distribution space than the region sampled over multiple trials (Fig. 1c). State-dependent decoding would therefore offer an improved encoding capacity, compared to the classical fixed-state decoder interpretation.

@@

Characterising noise correlations is complicated by the heterogeneity of sample response distribution structure and long-timescale changes 
% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
% Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also 
which, could cause correlative structure in the stimulus-response distribution (Fig. X), and must be controlled for when characterising `adaptation-independent' noise correlations.

@@

Using a large data set of spike sorted recordings in anesthetized rat barrel cortex, this paper aimed to precisely characterise the nature of short-timescale spike time noise correlations for first-spike patterns, by controlling for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
% towards estimating the potential encoding capacity of a state-dependent spike time encoding and determining the dimensionality of such an underlying latent state. The classical `stimulus-dependency' of noise correlations is left to future studies. 
A previous study analysing the same data set found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}. Another study analysing the data set could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}. 
% The spike time noise correlations uncovered in this study 
% offer a new opportunity to study population-wide modulation by low-dimensional latent states and 
% controls for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
% and was confirmed to occur for 7 out of 8 neuron groups in this study.

@@

Interestingly, many of the correlated clusters had non-unitary correlation angles. 
Importantly, many clusters remained correlated with non-unitary gradient after controlling for longer-timescale neural adaptation,
% using time series analysis, 
indicative of short-timescale fluctuating state, which varies the latency of different neurons at a different rate.
% giving evidence of precise non-adaptive noise correlations
% Such non-unitary spike time correlation angles cannot be 
This cannot explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle).

@@

The characterisation of adaptation-independent clustered spike-timing noise correlations is useful under the fixed-state decoder interpretations. More interestingly, these noise correlations open the possibility of state-dependent spike time coding. Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaptation-controlled correlated cluster, giving a prediction of the capacity of state-dependent encoding, which surpasses previously proposed coding forms.

Adaptation-independent non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers. This offers suggestive evidence of a low-dimensional fluctuating state affecting the spike times of spatially distanced neurons, which in turn could modulate downstream neurons and cause them to respond in a state-dependent manner. 
% The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. 
Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by low trial overlap between clusters. This paper provides
% in extracting and characterising adaptation-independent correlated clusters provides 
a useful step towards investigating this and shows that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.
% confirming or rejecting state-dependent spike-time coding.
